---
title: "imogen_analyses_comparisons"
author: "Imogen"
date: "2023-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up =  import packages and data

```{r}
#!/bin/R
### Install all packages
library(readr)
library(ggpicrust2)
library(tibble)
library(tidyverse)
library(ggprism)
library(patchwork)
library(ggh4x)
library(apeglm)

#read abundance data
abundance_file <- "picrust2_out_pipeline/pathways_out/path_abun_unstrat.tsv"
abundance_data <- read_delim(abundance_file, delim = "\t", col_names = TRUE, trim_ws = TRUE) %>% as.data.frame()
#read metadata
metadata <- read_delim("../colombia/metadata_categorized_CL.txt",delim = "\t",escape_double = FALSE,trim_ws = TRUE,show_col_types = FALSE)

#isolate smoking and nonsmoking from abundance data and metadata
metadata_smoking <- filter(metadata, metadata$smoker=="Yes")
metadata_nonsmoking <- filter(metadata, metadata$smoker=="No")

abundance_data_smoking <- select(abundance_data, metadata_smoking$`#SampleID`)
abundance_data_smoking$pathway = abundance_data$pathway
abundance_data_smoking <- abundance_data_smoking %>% select(pathway, everything())

abundance_data_nonsmoking <- select(abundance_data, metadata_nonsmoking$`#SampleID`)
abundance_data_nonsmoking$pathway = abundance_data$pathway
abundance_data_nonsmoking <- abundance_data_nonsmoking %>% select(pathway, everything())

### Load phyloseq object and subset according to smoker status
load("../DESeq module/phyloseq_object_final.RData") 
phylo_smoking <-  subset_samples(phyloseq_object_final, smoker == "Yes")
phylo_nonsmoking <-  subset_samples(phyloseq_object_final, smoker == "No")
taxa_info = as.data.frame(tax_table(phylo_smoking))
taxa_info$ASV = rownames(taxa_info)

```
# ALPHA DIVERSITY ANALYSIS


#BETA DIVERSITY ANALYSIS


# DESEQ2 ANALYSIS
## Processing phyloseq object into DESeq2 (smoking)
```{r, message=FALSE}
### Tidy up phyloseq object, relevel to set low LDL as baseline
phyloseq_object_final_plus1 <- transform_sample_counts(phylo_smoking, function(x) x+1)
phyloseq_object_final_deseq <- phyloseq_to_deseq2(phyloseq_object_final_plus1, ~ relevel(LDL_category, "low")) 
phyloseq_final <- DESeq(phyloseq_object_final_deseq)
### Ensure results is the healthy group
res <- results(phyloseq_final, tidy=TRUE)
colnames(res)[1] = "ASV"
res_with_taxa = inner_join(taxa_info,res, by = "ASV" )

### Get significant hits
res_sig = res_with_taxa %>% filter( padj<0.01 & abs(log2FoldChange)>2)

### count and report ASVs significantly up/down reg in smokers 
upregulated_count <- sum(res_sig$log2FoldChange > 0)
downregulated_count <- sum(res_sig$log2FoldChange < 0)
cat("Number of upregulated smoking ASVs:", upregulated_count, "\n")
cat("Number of downregulated smoking ASVs:", downregulated_count, "\n")

### Order results by log2FC
res_sig <- res_sig[order(res_sig$log2FoldChange),]
### Function to combine p values
combine_pvalues = function(p){return(1-pchisq(-2*sum(log(p),na.rm=T),2*sum(!is.na(p))))}

### Translating from ASV to genus level by merging the data for rows that have the same genus
### Need to take the average for log2FC and combined the Pvalues
res_genus_combined = res_sig %>%
  group_by(Genus) %>%
  summarize(log2FoldChange_avg = mean(log2FoldChange), pvalues =  combine_pvalues(pvalue)) %>%
  na.omit()
res_genus_combined <- res_genus_combined[order(res_genus_combined$log2FoldChange_avg),]

```
## Using DESeq2 data to produce barplots + volcano plots (smoking) @IMOGEN EDIT
```{r}
sighits_smok = ggplot(data = res_genus_combined, aes(x= log2FoldChange_avg,y=reorder(Genus, -(as.numeric(log2FoldChange_avg))), fill = pvalues))+
  geom_bar(stat = "identity") +
  theme_bw()+
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA)
### Save barplot
# ggsave("smoking_LDL_phyloseq_DeSeq.png", sighits_smok)

vol_plot_smok <- res %>%
  mutate(significant = padj<0.01 & abs(log2FoldChange)>2) %>% #new column in results table 
  ggplot() +
  geom_point(aes(x=log2FoldChange, y=-log10(padj), col=significant)) + 
  labs(x = "Log2 Fold Change", y = "-Log10 (padj)")
vol_plot
### Save volcano plot
# ggsave(filename="vol_plot_smokingLDL.png",vol_plot_smok)
```
## Processing phyloseq object into DESeq2 (NONsmoking)
```{r}
### Tidy up phyloseq object, relevel to set low LDL as baseline
phyloseq_object_final_plus1 <- transform_sample_counts(phylo_nonsmoking, function(x) x+1)
phyloseq_object_final_deseq <- phyloseq_to_deseq2(phyloseq_object_final_plus1, ~ relevel(LDL_category, "low")) 
phyloseq_final <- DESeq(phyloseq_object_final_deseq)
### Ensure results is the healthy group
res_nonsmoke <- results(phyloseq_final, tidy=TRUE)
colnames(res_nonsmoke)[1] = "ASV"
res_nonsmoke_taxa = inner_join(taxa_info,res_nonsmoke, by = "ASV" )

### Get significant hits
res_nonsmoke_sig = res_nonsmoke_taxa %>% filter( padj<0.01 & abs(log2FoldChange)>2)

### count and report ASVs significantly up/down reg in smokers 
upregulated_count <- sum(res_nonsmoke_sig$log2FoldChange > 0)
downregulated_count <- sum(res_nonsmoke_sig$log2FoldChange < 0)
cat("Number of upregulated nonsmoking ASVs:", upregulated_count, "\n")
cat("Number of downregulated nonsmoking ASVs:", downregulated_count, "\n")

### Order results by log2FC
res_nonsmoke_sig <- res_nonsmoke_sig[order(res_nonsmoke_sig$log2FoldChange),]

### Translating from ASV to genus level by merging the data for rows that have the same genus
### Need to take the average for log2FC and combined the Pvalues
res_nonsmoke_genus_combined = res_nonsmoke_sig %>%
  group_by(Genus) %>%
  summarize(log2FoldChange_avg = mean(log2FoldChange), pvalues =  combine_pvalues(pvalue)) %>%
  na.omit()
res_nonsmoke_genus_combined <- res_nonsmoke_genus_combined[order(res_nonsmoke_genus_combined$log2FoldChange_avg),]

```
## Using DESeq2 data to produce barplots + volcano plots (NONsmoking) @IMOGEN EDIT
```{r}
sighits_nonsmok = ggplot(data = res_nonsmoke_genus_combined, aes(x= log2FoldChange_avg,y=reorder(Genus, -(as.numeric(log2FoldChange_avg))), fill = pvalues))+
  geom_bar(stat = "identity") +
  theme_bw()+
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA)
### Save barplot
ggsave("nonsmoking_LDL_phyloseq_DeSeq.png", sighits_nonsmok)

vol_plot_nonsmok <- res_nonsmoke %>%
  mutate(significant = padj<0.01 & abs(log2FoldChange)>2) %>% #new column in results table 
  ggplot() +
  geom_point(aes(x=log2FoldChange, y=-log10(padj), col=significant)) + 
  labs(x = "Log2 Fold Change", y = "-Log10 (padj)")
vol_plot
### Save volcano plot
ggsave(filename="vol_plot_nonsmokingLDL.png",vol_plot_nonsmok)
```

# PICRUST ANALYSIS - Setup
```{r, message=FALSE}
### PICRUSt2 is pickier about the formatting of the data, so first run this block to ensure everything is as needed.
###Load a custom function by Chris to provide all DESeq2 results
source("../picrust_processing/DESeq2_function.R")
###Removing all columns that include NA's for the pathways_daa() function
metadata = metadata[ , colSums(is.na(metadata)) == 0]
###Filter your metadata as needed to look at specific comparisons
metadata_smoking = metadata %>%filter(smoker == "Yes")
metadata_nonsmoking = metadata %>%filter(smoker == "No")

###Filtering the abundance table to only include samples that are in the filtered metadata
sample_names_smoking = metadata_smoking$`#SampleID`
sample_names_smoking = append(sample_names_smoking, "pathway")
abundance_data_smoking = abundance_data[, colnames(abundance_data) %in% sample_names_smoking] 

sample_names_nonsmoking = metadata_nonsmoking$`#SampleID`
sample_names_nonsmoking = append(sample_names_nonsmoking, "pathway")
abundance_data_nonsmoking = abundance_data[, colnames(abundance_data) %in% sample_names_nonsmoking] 

###Removing individuals with no data that caused a problem for pathways_daa()
abundance_data_smoking =  abundance_data_smoking[, colSums(abundance_data_smoking != 0) > 0]
abundance_data_nonsmoking =  abundance_data_nonsmoking[, colSums(abundance_data_nonsmoking != 0) > 0]

###Filtering out columns that represent a total abundance < 10000
###Filtering out rows (pathways) that have a total count < 100
abundance_data_smoking = abundance_data_smoking[,colSums(abundance_data_smoking[,-1]) > 10000]
abundance_data_smoking = abundance_data_smoking[rowSums(abundance_data_smoking[,-1]) > 100,]
abundance_data_nonsmoking = abundance_data_nonsmoking[,colSums(abundance_data_nonsmoking[,-1]) > 10000]
abundance_data_nonsmoking = abundance_data_nonsmoking[rowSums(abundance_data_nonsmoking[,-1]) > 100,]

###Ensuring the rownames for the abundance_data_filtered is empty. This is required for their functions to run.
rownames(abundance_data_smoking) = NULL
rownames(abundance_data_nonsmoking) = NULL

###Verify samples in metadata match samples in abundance_data
abun_samples_smoking = rownames(t(abundance_data_smoking[,-1])) #Get list of sample names in  newly filtered abundance data
metadata_smoking = metadata_smoking[metadata_smoking$`#SampleID` %in% abun_samples_smoking,] #Ensure filtered metadata only includes these samples
abun_samples_nonsmoking = rownames(t(abundance_data_nonsmoking[,-1]))
metadata_nonsmoking = metadata_nonsmoking[metadata_nonsmoking$`#SampleID` %in% abun_samples_nonsmoking,]

### Filter data to remove 0 standard deviation
row_sds_nonsmok <- abundance_data_nonsmoking %>% column_to_rownames("pathway") %>% apply(1, sd)
filtered_LDL_nonsmok <- abundance_data_nonsmoking[row_sds_nonsmok != 0, ]
row.names(filtered_LDL_nonsmok) <- NULL
row_sds_smok <- abundance_data_smoking %>% column_to_rownames("pathway") %>% apply(1, sd)
filtered_LDL_smok <- abundance_data_smoking[row_sds_smok != 0, ]
row.names(filtered_LDL_smok) <- NULL

# Perform pathway DAA using DESeq2 method
metacyc_daa_results_df_smoking <- pathway_daa(abundance = abundance_data_smoking %>% column_to_rownames("pathway"), metadata = metadata_smoking, group = "LDL_category", daa_method = "DESeq2")
# Annotate MetaCyc pathway results without KO to KEGG conversion. This line provides a useful description for your pathway accessions. 
metacyc_daa_annotated_results_df_smoking <- pathway_annotation(pathway = "MetaCyc", daa_results_df = metacyc_daa_results_df_smoking, ko_to_kegg = FALSE)

metacyc_daa_results_df_nonsmoking <- pathway_daa(abundance = abundance_data_nonsmoking %>% column_to_rownames("pathway"), metadata = metadata_nonsmoking, group = "LDL_category", daa_method = "DESeq2")
metacyc_daa_annotated_results_df_nonsmoking <- pathway_annotation(pathway = "MetaCyc", daa_results_df = metacyc_daa_results_df_nonsmoking, ko_to_kegg = FALSE)

#Generating the full DESeq2 results dataframe
#Here is where we can see HOW up or downregulated a certain pathway is.
#The reference group is selected alphabetically. In this case, Dogs are the reference and anything with a positive log2foldchange means it is higher in humans.
smoking_res = DEseq2_function(abundance_data_smoking, metadata_smoking)
smoking_res$feature = rownames(smoking_res)
nonsmoking_res = DEseq2_function(abundance_data_nonsmoking, metadata_nonsmoking)
nonsmoking_res$feature = rownames(nonsmoking_res)

#Joining description
smoking_res_desc = inner_join(smoking_res, metacyc_daa_annotated_results_df_smoking, by = "feature")
smoking_res_desc = smoking_res_desc[, -c(8:13)]
nonsmoking_res_desc = inner_join(nonsmoking_res, metacyc_daa_annotated_results_df_nonsmoking, by = "feature")
nonsmoking_res_desc = nonsmoking_res_desc[, -c(8:13)]

#Filter by p-value <0.05
smoking_res_desc_sig = smoking_res_desc %>% filter(pvalue <0.05)
nonsmoking_res_desc_sig = nonsmoking_res_desc %>% filter(pvalue <0.05)
```
## Graph PICRUSt results in barplots, filtered for >1 log2 fold chance
```{r}
smoking_res_desc_sig <- smoking_res_desc_sig[order(smoking_res_desc_sig$log2FoldChange),]
nonsmoking_res_desc_sig <- nonsmoking_res_desc_sig[order(nonsmoking_res_desc_sig$log2FoldChange),]

###Filtering for >1 fold change
smoking_res_above1 <- smoking_res_desc_sig %>%
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>% arrange(log2FoldChange)
nonsmoking_res_above1 <- nonsmoking_res_desc_sig %>%
  filter(log2FoldChange > 1 | log2FoldChange < -1) %>% arrange(log2FoldChange)

### Barplot for smoking results
ggplot(data = smoking_res_above1, aes(x= log2FoldChange,y=reorder(description, -(as.numeric(log2FoldChange))), fill = pvalue))+ theme_bw()+
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA)+
  xlim(-0.5,2.5)+
  labs(x = "log2 Fold Change", y = "Bacterial Pathway")+
  ggtitle("LDL-Associated Pathway Changes in Smokers")

### Barplot for nonsmoking results
ggplot(data = nonsmoking_res_above1, aes(x= log2FoldChange,y=reorder(description, -(as.numeric(log2FoldChange))), fill = pvalue))+ theme_bw()+
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA)+
  xlim(-0.5,2.5)+
  labs(x = "log2 Fold Change", y = "Bacterial Pathway")+
  ggtitle("LDL-Associated Pathway Changes in Non-Smokers")

### Barplot for smoking + nonsmoking results combined
smoking_res_above1$smoking="S"
nonsmoking_res_above1$smoking="NS"
figure4_res_above1 <- rbind(smoking_res_above1, nonsmoking_res_above1)
figure4_res_above1$smoking = factor(figure4_res_above1$smoking, levels=c("S","NS"))

ggplot(data = figure4_res_above1, aes(x= log2FoldChange,y=reorder(description, -(as.numeric(log2FoldChange))), fill = pvalue))+ theme_bw()+
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "yellow", high = "red", na.value = NA)+
  xlim(-0.5,2.5)+
  labs(x = "Log2 Fold Change", y = "Bacterial Pathway")+
  facet_grid(rows = vars(smoking),scales ="free_y", space = "free_y")+
  theme(strip.text = element_text(
    size = 10, color = "black"), 
    axis.text.y = element_text(size = 18, face = "bold"), 
    legend.text = element_text(size = 12, face = "bold"))
```

## Using ggpicrust to create PCA plots, supplemental figure 4
```{r, message=FALSE}
PCA_smoking_all <- pathway_pca(abundance = abundance_data %>% column_to_rownames("pathway"), 
                         metadata = metadata, group = "smoker")+
                        ggtitle("PCA plot comparing smoking in all participants")
PCA_LDL_all <- pathway_pca(abundance = abundance_data %>% column_to_rownames("pathway"), 
                           metadata = metadata, group = "LDL_category")+
                        ggtitle("PCA plot comparing LDL in all participants")
PCA_nonsmok_LDL <- pathway_pca(abundance = filtered_LDL_nonsmok %>% column_to_rownames("pathway"), 
                               metadata = metadata_nonsmoking, group = "LDL_category")+
                                ggtitle("            Non-smokers")
PCA_smok_LDL <- pathway_pca(abundance = filtered_LDL_smok %>% column_to_rownames("pathway"), 
                            metadata = metadata_smoking, group = "LDL_category")+
                                ggtitle("            Smokers")
PCA_smoking_all
PCA_LDL_all
PCA_nonsmok_LDL
PCA_smok_LDL
```